---
title: "Crime Rates"
---

Follow along with the demo to learn how to calculate crime rates using R and the Tidyverse. We'll also introduce a new library, janitor, which helps make column names easier to work with. You should get started by loading the tidyverse in a new R block, then install and load janitor.

### Task 1: Load the tidyverse and install and load janitor

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)

```


### Task 2: Load the crime dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv

```{r}
crimes <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_annual_crime.csv") %>% clean_names()
```


### Task 3: Use janitor to clean up the column names



### Task 4: Calculate the total crime rate for each year (picking our unit!)

```{r}
crimes %>% mutate(total_rate = grand_total/population * 10000)
```


### Task 5: Calculate the violent and property crime rates for each year (same unit)

```{r}
crimes_with_rates <- crimes %>% mutate(violent_rate = violent_crime_total/population * 10000,property_rate = property_crime_total/population * 10000)
```


### Task 6: Describe our results

Write a few sentences about what you found in the data. What years had the highest crime rates? What years had the lowest crime rates? Does the pattern match your expectations?

2020 had the lowest crime rates, and 1980 had the highest crime rates. The patterns do match my expectations, as we learned in class about how America overall has gotten safer over time. The 1980s having the highest crime rates also makes sense with the high drug use at the time. I do think it is interesting that there was a lower population in the 70s and 80s, yet they had more crime. It makes me wonder if there has been more enforcement and that is why crime rates have changed, or if there is another factor.

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

