---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80
```{r}
set.seed(42)
population <- tibble(
  age = sample(18:80, 1000, replace=TRUE)
)
head(population)
```


### Task 3: Sample 100 rows of data from the dataset
```{r}
sample_100 <- population %>% 
    sample_n(100)

sample_50 <- population %>% 
    sample_n(50)

sample_10 <- population %>% 
    sample_n(10)
```


### Task 4: Calculate the mean age and standard deviation of the sample

```{r}
sample_100 %>% summarize(mean = mean(age), sd = sd(age))

sample_50 %>% summarize(mean = mean(age), sd = sd(age))

sample_10 %>% summarize(mean = mean(age), sd = sd(age))
```


### Task 5: Calculate the mean age and standard deviation of the population

```{r}
population %>% summarize(mean = mean(age), sd = sd(age))
```


### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

The mean and standard devitation stay pretty consistent for the sample sizes of 100 and 50, but they decrease for the mean by about 7 when looking at a sample size of 10. The standard devitation was fairly consistent, but the sample size of 100 and 50 had a a closer standard deviation. This emphasizes that a good size sample is needed to get a grasp on the results for a larger population, as the sample size can determine the accuracy of the data. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
