---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

I think that the assault shooting category is important, as it is a severe crime and it also has an interesting trend, with only 5 accounts in January, and 21 accounts in July and August. A good story would consist of this data and any factors that may contribute to the uptick in these kinds of assaults, such as regulation or social changes. Many of the categories have stayed fairly consistent, meaning that there are no clear outliers, and it may be interesting why those numbers have stayed the same and why any policies put in place to combat them are not effective. Homicide and crimes considered more dangerous may be considered more important, but ultimately, I believe any crime trends are newsworthy and warrant a trend explanation.
The assualt shooting category may make it difficult to summarize the data due to the large variety in data, so a report would not be able to say that crime has stayed fairly consistent overall.
### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> 
  mutate(total_incidents = `Jan` + `Feb` + `Mar` + `Apr` + `May` + `Jun` + `Jul` + `Aug` + `Sep` + `Oct` + `Nov` + `Dec`) |> 
  mutate(average_incidents = total_incidents / 12)
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The categories with the highest averages are Accident, Auto Stolen, Theft from Auto and Accident with Impound. The categories with the lowest averages are Homicide, B&E Other, Sex Offense and Assault Weapon. This means that the highest amount of incidents have to do with accidents and cars, and there are less serious incidents, such as homicide and sex offenses. I have questions about the auto stolen and the theft from auto, as I feel like these incidents are not as common as the data explains. I have questions about what models of cars they are taking the most of, if the victims were locking their cars, if there is evidence of one suspect or multiple, and if the methods are consistent throughout incidents.I have questions about other data points, but the trend of car incidents stuck out to me. 

### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> 
  mutate(june_change = (Jun - average_incidents) / average_incidents * 100) |> 
  mutate(july_change = (Jul - average_incidents) / average_incidents * 100) |> 
  mutate(august_change = (Aug - average_incidents) / average_incidents * 100)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

I could not figure out the code for the previous answer, but based on my observations of the data, crime does not seem that much more likely to occur in the summer. There are some categories that had crime in higher numbers during the summer months, such as "Auto Stolen," but most of the categories had about the same or only slightly higher levels. I do not think that this theory can be generalized, as it may depend on the location and the type of crime someone is discussing. 

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Assault shootings at an extreme low in January in Prince George's County, compared numbers more than four times as large in July and August. 

January had five total assault shootings in the county, while July and August each had 21 incidents. An average of about 16 assault shootings take place each month in Prince George's County, making five a notably low number for the month of January. There were a total of 188 assualt shootings in the year. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
