---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
library(dplyr)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The dataframe displays the salaries of people in the White House, ranging from $0 to $251,258. The salaries that are zero dollars stood out to me, as I was confused as to why those people make no money. They also all have different roles, and some of their titles include "Senior Director." I would assume that someone with that title in their name would make more money, as some of the other higher paid roles have that title. The status column also has two categories: employee and detailee. I had to look up what a detailee is-it is someone who is assigned to a different department for a period of time. I think that the $0 salaries may cause some problems for analysis, as I am not sure if this is volunteer work or a mess up in the data/organization. The rest of the salary values seem valid for analysis.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries %>% summarize(mean_salary=mean(salary))
wh_salaries %>% summarize(median_salary=median(salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries %>% summarize(mean_salary=mean(salary))
wh_salaries %>% summarize(median_salary=median(salary))
wh_salaries %>% summarize(min_salary=min(salary))
wh_salaries %>% summarize(max_salary=max(salary))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and median salaries are similar, with only about $3,000 between them. This is interesting because there were 11 rows that had $0 for their salary, which I thought would have skewed the data more. I believe these values still brought the mean down, as it was less than the median. Since the mean and median are similar, but not the same, the data is not perfectly symmetrical, but not completely skewed. I think that the multiple $0 is an outlier, as the next highest value is $20,000. Since there are multiple of them, it is not an extreme outlier. The minimum and maximum tells us the large range of the data-$251,258. This is also the max value, since the min value is 0. Overall, the data tells us the range in salaries, what the average is, and  begs the question of why the low is $0.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

Based on this data, I think the best story idea would be to investigate why 11 people have a salary of $0. This would mean doing a deep dive into their role, seeing how it compares to people in similar roles or similar titles, seeing if I could get in contact with some of the people in the roles and investigating if there is a reason they are working for free. I am also wondering if it was a mistake in the data, as some of these people have titles that bring a reasonable assumption of power, so their lack of pay is questionable. If the people I reach out to do not answer, I could look into using FOIA to get some more data or information.
