---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities %>% mutate(violent_rate_2019 = violent2019/pop2019 * 1000, property_rate_2019 = property2019/pop2019 * 1000, violent_rate_2020 = violent2020/pop2020 * 1000, property_rate_2020 = property2020/pop2020 * 1000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

Baltimore and Ocean City had the highest amount of violent crimes per 1,000 people in both 2020 and 2019. In 2019, Baltimore had 18 violent crimes per 1,000 people, while Ocean City had 13 violent crimes per 1,000 people. In 2020, the two cities switched positions, with Ocean City having 23 violent crimes per 1,000 people, and Baltimore having 15. Ocean City also had the highest amount of property crimes in 2019 and 2020, going from 103 to 121 per 1,000 people. Overall, the data shows that Ocean City has a high amount of crimes, as it had the most crimes for 3/4 categories observed.
It is also significant that a large amount of cities had no violent crimes, and many of these cities are in Montgomery county. The same is not necessarily true about Montgomery county and property crimes- there was more variation in counties with no property crimes compared to violent crimes.
### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates %>% mutate(pct_violent = violent_rate_2020 - violent_rate_2019/violent_rate_2019 *100, pct_property = property_rate_2020 - property_rate_2019/property_rate_2019 *100 )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

One outlier that sticks out to me is the 21% increase in property crime in Ocean City. All other percent changes were significant decreases, except for property crime in Ocean City. 
Aside from the outlier, all of the data had decreases ranging from 100% to about 40%.I believe that this is attributed to the pandemic, as there was not as much opportunity for crime. I think that this is a significant pattern, but not more interesitng than the outlier. Most people would predict that crime would decrease due to the pandemic, but the property crime increase in Ocean City is something that would need to be investigated.
### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Ocean City saw a 21% increase in property crimes in 2020, while all other Maryland cities saw significant decreases.

While Baltimore City saw a 69% decrease in property crime in 2020 and Oakland saw a 44% decrease, Ocean City's property crimes rose form 720 to 840 incidents. The Covid-19 pandemic lead to many shut downs and is partially credible for the overall lower crime rates, yet Ocean City defies that idea by having the only property crime increase in Maryland.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
