---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
songs %>% summarize(mean = mean(danceability), sd = sd(danceability))
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
songs %>% 
  ggplot() +
geom_histogram(aes(x = danceability))
```

### Task 5: Add a line to the histogram for the mean danceability score

Write code to add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
songs %>% 
  ggplot() +
geom_histogram(aes(x = danceability)) +
geom_vline(aes(xintercept = mean(danceability)), color = "blue", linetype= "dashed", size=1)
```

### Task 6: Describe our results

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and distribution tell us about the songs in the dataset?

The distribution is not exactly normal, as it is skewed to the right since there is more frequency of numbers on the right, or their bar graphs are consistently higher. This means the same amount of songs have a danceability ranging from 0-0.67 as the amount of songs with danceability ranging from 0.68-1. Therefore, the data is skewed to the right. The mean and distribution show that the songs in the dataset have a general high danceability, as a majority have a score over 0.5. The distribution also shows us what values impact the mean, such as the significantly high bar that is directly to the right of the mean. This value, having a higher count than most, must have had an important impact on the mean, and means that many songs have a danceability around 0.7.

### Task 7: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

