---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
songs %>% summarize(mean = mean(danceability), sd = sd(danceability))
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
songs %>% 
  ggplot() +
geom_histogram(aes(x = danceability))
```

### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
songs %>%  
ggplot() +
  geom_histogram(aes(x = danceability)) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1)
```


### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
songs %>%  
ggplot() +
  geom_histogram(aes(x = danceability)) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) - sd(danceability)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) + sd(danceability)), color = "blue", linetype = "dashed", size = 1)
```

### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

The distribution of the data is skewed to the right, as the distribution does not show a perfectly normal or symmetrical bell curve. The mean shows the average danceability of the songs, which is 0.67, and that many of the songs in the dataset would be considered danceable. Since the distribution is not normal, the  standard deviation percentage rule does not apply. Therefore, it is safe to assume that more of the data falls within one standard deviation above the mean, as opposed to being equal with one standard deviation below the mean. This is evident from the histogram, and the cluster of data to the right of the mean. 

Based on this data, I would pitch a story about how many songs are considered danceable and cite data about the cluster of data within one standard deviation above the mean. It would also be easy to pull songs that are within the IQR of the data to further elaborate on my point. 
Another story idea would be the most "danceable" songs of the year and to identify how they compare with the average danceability score for that year. 
### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
