---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.

```{r}
sample5 <- data %>% sample_n(5)

sample10 <- data %>% sample_n(10)
```


### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.

```{r}
sample5 %>%  summarize(mean = mean(red), sd= sd(red))

sample10 %>%  summarize(mean = mean(red), sd=sd(red))
```


### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4

```{r}
data %>%  summarize(mean = mean(red), sd = sd(red))
```


### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?

The three means are similar, as the population's mean of red M&Ms is about 73, while the sample of 5 rows' mean is 78 and the sample of 10 rows' mean is 73. The sample of 10 and the entire population have about the same rounded mean, and are only a few decimals away from each other. The sample of five rows is slightly farther away. 
The standard deviation of the samples are similar, with both being about 21. The populations' standard deviation is 16. This is not too far off, but still a significant difference to note. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

The data shows that a larger sample size will be more accurate to reflect the data in the population. The data also showed that each sample will be different, as different parts of the data are being randomly selected during the sampling process. While sampling will not get you perfect results, as shown by the slight difference in the 10 row sample vs. the population, having a large sample will generally get the data closer to the actual population data.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
