---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: "Nicole Pilsbury"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

The independent variable that is most correlated with voter turnout is the percentage of white people in the county. This is followed by the median age and the income of the county. The percentage of unemployed people in the county is very negatively correlated with voter turnout.

The percentage of people with bachelor's degrees is also decently correlated with the income of the community. The percentage of white people is negatively correlated with the overall population.

The median age and the percent of people with bachelors degrees are not correlated at all.

The percentage of unemployed people has a negative linear relationship with voter turnout, as the data points closely follow the line of best fit. The median age and percent of people with bachelors degrees have a somewhat linear relationship, but there are some outliers.

The population, percent of white people and income have a decent amount of outliers that make me question a linear relationship between the variables. Many of the data points do not align with the line of best fit.  

```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

This model predicts about 80% of the data, which I think means that the model does a fairly good job at predicting voter turnout. It also has a low p-value, meaning that the data is statistically significant.
Income, median age and percent white all had low p-values and statistically significant data. All of the coefficients are very low, with all of them needing scientific notation to be written out. The largest coefficient is income, followed by population (with a negative coefficient).
I think that this is a useful model because it explains that 80% of the data can be explained, and it also shows us which of the independent variables are statistically significant.
```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

I chose to center my model around median age, income and percent white since they were the statistically significant variables in the full model.
Model 3, which considers all three of the listed independent variables, was tied with the ful model for an adjusted R-squared value of about 80%. Model 3 is above the full model, which I am assuming means it has a higher value.
Model 2, which took income and percent white into account, took had a higher adjusted R-squared value by about 30 percentage points compared to model 4, which only took median age and percent white into account.

```{r}
# Model 2
model2 <- lm(turnout ~ income + pct_white, data = md_turnout)
summary(model2)

# Model 3
model3 <- lm(turnout ~ pct_white + median_age + income, data = md_turnout)
summary(model3)

# Model 4
model4 <- lm(turnout ~ median_age + pct_white, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

You could use this analysis to cover an election story by saying that a county's voter turnout can be predicted most accurately when taking income, age and the percentage of white people into account. I think people would understand if the story was written in a way that explained how 80% of the data can be accurately predicted by these three factors.
There would also have to be addition of quotes and specific information about how these factors contribute to turnout (increase/decrease).
