---
title: "Prince George's County Car Theft Spike Analysis"
author: "Nicole Pilsbury"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Introduction

This analysis examines the impact of a significant spike in car theft rates in 2023 in Prince George's County, and whether that spike has led to a sustained higher rate of car theft in the county. We'll use Analysis of Variance (ANOVA) to determine if theft rates differ significantly between pre-spike (before 2023), spike (2023), and post-spike (2024-present) periods.

## Data Preparation and Exploration

### Loading Libraries and Data

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(lubridate)
```

```{r load-data}
# Import data
theft_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month17_25.csv")

# Look at the first few rows
head(theft_data)

# Get a summary of the data
summary(theft_data)
```


### Get the data how we want it and calculate the rates

**Task 1** Replace REPLACE_ME with the variables you need to calculate the theft_rate (2 points)

```{r process-data}
# Calculate theft rate and create time period variable
theft_data <- theft_data |>
  mutate(
    date = as.Date(month),
    year = year(date),
    month_number = month(date),
    theft_rate = (total / population) * 100000,
    time_period = case_when(
      year < 2023 ~ "Pre-spike",
      year == 2023 ~ "Spike year",
      year > 2023 ~ "Post-spike"
    )
  )

# Create summary by time period
period_summary <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_thefts = mean(total),
    mean_rate = mean(theft_rate),
    median_rate = median(theft_rate),
    sd_rate = sd(theft_rate),
    n = n(),
    .groups = "drop"
  )

# Display the summary
knitr::kable(period_summary, digits = 2,
             caption = "Summary of Car Theft Rates by Time Period")
```

**Task 2** Describe the results of calculating the rates for each period. What do they suggest? (3 points)

The results suggest that the rate of car thefts per 100,000 people has increased compared to the pre-spike time period. The spike year had the largest rate of about 60, while the pre-spike period had a rate of about 23 and the post-spike rate was about 43. This data looks to be proving that the rate had a sustained increased after the spike year. 

## Visualizing the Spike Pattern

### Time Series Plot

```{r time-series}
# Time series plot showing the spike pattern
ggplot(theft_data, aes(x = date, y = theft_rate)) +
  geom_line() +
  geom_smooth(method = "loess", span = 0.2) +
  geom_vline(xintercept = as.Date("2023-01-01"), linetype = "dashed", color = "red") +
  geom_vline(xintercept = as.Date("2024-01-01"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("2023-07-01"), y = max(theft_data$theft_rate) * 0.9, 
           label = "Spike Year", color = "red") +
  labs(title = "Monthly Car Theft Rates (2017-2025) with Spike Highlighted",
       x = "Date", y = "Thefts per 100,000 Population") +
  theme_minimal()
```

**Task 3** Describe the output of this chart and then give it a better, more active title that provides the lede (3 points)

The output of this chart visually shows that the rate of car thefts increased after the spike year when comapred to the pre-spike period, as the theft rate is consisently higher on the chart. The chart does show a downward curve towards the end, which could lead to more investigation, but it is still higher than the pre-spike period data. 

Active Title: Monthly Car Theft Rates Sustain Increase After Spike in 2023


Now let's take a look at average theft rates.

### Yearly Average Theft Rates

```{r yearly-means}
# Calculate yearly means
yearly_means <- theft_data |>
  group_by(year) |>
  summarize(
    mean_rate = mean(theft_rate),
    .groups = "drop"
  )

# Plot yearly means
ggplot(yearly_means, aes(x = year, y = mean_rate)) +
  geom_line(size = 1) +
  geom_point(size = 3, color = "blue") +
  geom_point(data = yearly_means |> filter(year == 2023), 
             aes(x = year, y = mean_rate), size = 4, color = "red") +
  labs(title = "Average Annual Car Theft Rates",
       x = "Year", y = "Average Thefts per 100,000 Population") +
  theme_minimal() +
  scale_x_continuous(breaks = 2017:2025)
```

## One-way ANOVA Analysis

### Hypothesis Formulation

For our ANOVA test, we'll compare the mean theft rates between the three time periods. First we need to create our hypotheses.

**Task 4** Fill in the hypotheses (2 points)

- **Null Hypothesis (H0)**: The spike year did not lead to a sustained higher rate of car thefts in PG county. 

- **Alternative Hypothesis (H1)**: The spike year led to a sustained higher rate of car thefts in PG county. 

### One-way ANOVA Test

```{r anova-test}
# Perform one-way ANOVA on theft rates by time period
period_anova <- aov(theft_rate ~ time_period, data = theft_data)
summary(period_anova)
```

**Task 5** Describe the result: are there real differences in the average rates between the time periods? Is this result significant? (5 points)

The large f-value means that there are differences between the groups (time periods), not just within them. This also means that the differences between the groups are larger than those within. The p-value is also low, meaning that we can reject our null hypothesis and accept the alternative hypothesis of the spike year leading to a sustained higher rate of car thefts. Both the f-value and the p-value confirm our alternative hypothesis and show that the result is significant. 

## Compare the differences between specific periods

Run our post-anova Tukey test and examine the results. 

```{r posthoc}
# Tukey's HSD test
tukey_results <- TukeyHSD(period_anova)
print(tukey_results)
```


```{r period-means}
# Visualize time period means with confidence intervals
period_rates <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_rate = mean(theft_rate),
    se = sd(theft_rate) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    lower_ci = mean_rate - qt(0.975, n() - 1) * se,
    upper_ci = mean_rate + qt(0.975, n() - 1) * se
  )

# Ensure time period is ordered correctly
period_rates$time_period <- factor(period_rates$time_period, 
                                   levels = c("Pre-spike", "Spike year", "Post-spike"))

ggplot(period_rates, aes(x = time_period, y = mean_rate, fill = time_period)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  labs(title = "Car Theft Rates by Time Period with 95% Confidence Intervals",
       x = "", y = "Mean Theft Rate per 100,000") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Conclusion

This analysis has examined whether there are significant differences in car theft rates between the pre-spike period (before 2023), the spike year (2023), and the post-spike period (after 2023).

**Task 6** Describe the results of the Tukey's test and chart showing confidence intervals. Is there a significant difference between the three periods? In particular, is the difference between the pre-spike and post-spike periods real and significant? Write a headline and lede reporting these findings, focusing on the pre- and post-spike periods. (5 points)

Answer: There is a significant difference between teh three periods, as the spike year is greater than both the pre and post spike categories. The post-spike is also greater than the pre-spike, which means that there was a sustained higher rate after the spike year. The p-value and the f-value show that the results are significant. The box plot helps to visualize the higher theft rate in the post-spike time compared to the pre-spike time.

Headline: Car thefts in Prince George's County increase after 2023 spike

Lede: Forty-three out of 100,000 Prince George's County residents are at risk of car theft, compared to 23 out of 100,000 before the car theft spike in 2023. 
