---
title: "HW11: Education Level and Survey Weighting"
author: "Nicole Pilsbury"
---

```{r setup, include=FALSE}
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Introduction

In this assignment, you'll examine how survey weighting affects our understanding of voting intentions across different education levels, using the data we examined in class. You'll calculate both weighted and unweighted statistics, create visualizations, and reflect on the implications for reporting.

## The Data

The nonvoters dataset contains survey responses about voting intentions and behaviors from a national survey. The survey was conducted prior to an election and includes demographic information like education level.

```{r}
# Load the dataset
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

# Take a quick look at the data structure
glimpse(nonvoters_data)
```

### Key Variables

- `weight`: Survey weight assigned to each respondent
- `Q21`: Voting intention (1 = Yes, 2 = No, 3 = Unsure/Undecided)
- `educ`: Education level (College, Some college, High school or less)

## Task 1: Education Distribution

First, let's examine the distribution of education levels in our sample. Replace "REPLACE_ME" with the correct variable for education level.

```{r}

education_distribution <- nonvoters_data |>
  count(educ) |>
  mutate(percentage = n / sum(n) * 100) |>
  kable(digits = 1, col.names = c("Education Level", "Count", "Percentage (%)"))

education_distribution
```

## Task 2: Reflection Question

Why might education levels in survey samples often differ from the general population? What factors might cause certain education groups to be over or underrepresented?

Education levels might differ from the general population because it may be more difficult to get a hold of those who have less than a college level education. This may be because of a lack of access to technology or information, or they may be busy with jobs to support themselves. Those will a college education tend to be wealthier, and have more access to technology, information, and time to take surveys. Therefore, those that are less educated are less likely to respond to surveys. 


## Task 3: Unweighted Analysis by Education

Now, let's calculate unweighted voting intentions by education level. This is what we would report if we didn't apply any weighting to our sample.

```{r}
# Calculate unweighted voting intentions by education
unweighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Count responses
  summarize(count = n(), .groups = "drop_last") |>
  # Calculate percentages
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

# Create a more readable format with voting intentions as columns
unweighted_educ_summary <- unweighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

kable(unweighted_educ_summary, digits = 1, caption = "Unweighted Voting Intentions by Education Level")
```

## Task 4: Reflection Question

Based just on this unweighted analysis, what headline might you write for a news story about education and voting intentions?

College educated citizens most likely to vote in upcoming election

## Task 5: Weighted Analysis by Education

Next, let's apply survey weights to see how this changes our results. Instead of just counting responses, we'll sum the weights for each group. Replace "REPLACE_ME" with the appropriate weight variable

```{r weighted-by-education}

weighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Sum the weights instead of counting
  summarize(weighted_count = sum(weight), .groups = "drop_last") |>
  # Calculate weighted percentages
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

# Create a more readable format
weighted_educ_summary <- weighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

kable(weighted_educ_summary, digits = 1, caption = "Weighted Voting Intentions by Education Level")
```

## Task 6: Reflection Questions

1. How did the percentages change after applying weights? Which education group showed the biggest changes?

The percentages only changed by a few percentage points, with the high school or less group changing the most across all three categories. The other two groups did not see much change. 

2. Why might the weighted results be considered more accurate than the unweighted results?

The weighted results might be more accurate than the unweighted because they accurately match the demographics of the population. They ensure that no group is over or under represented. 

## Task 7: Comparison of Weighted vs. Unweighted Results

Let's create a direct comparison table to see the differences more clearly.

```{r}

comparison <- unweighted_educ_summary |>
  inner_join(weighted_educ_summary, by = "educ", suffix = c("_unweighted", "_weighted")) |>
  mutate(
    # Calculate the differences between weighted and unweighted percentages
    yes_diff = `Yes (%)_weighted` - `Yes (%)_unweighted`,
    no_diff = `No (%)_weighted` - `No (%)_unweighted`,
    unsure_diff = `Unsure (%)_weighted` - `Unsure (%)_unweighted`
  ) |>
  # Select just the columns we want to display
  select(educ, yes_diff, no_diff, unsure_diff) |>
  rename(
    "Education Level" = educ,
    "Yes (% point diff)" = yes_diff,
    "No (% point diff)" = no_diff,
    "Unsure (% point diff)" = unsure_diff
  )

kable(comparison, digits = 1, caption = "Difference Between Weighted and Unweighted Results (percentage points)")
```

## Task 8: Reflection Question

Which education group shows the largest differences between weighted and unweighted results?

The high school or less group showed the largest difference.

## Task 9: Visualization

Visualizations can help us see the differences more clearly. Let's create a bar chart comparing weighted and unweighted "Yes" responses by education level. Replace "REPLACE_ME" with the correct variable name

```{r}
educ_viz_data <- bind_rows(
  # Unweighted data
  unweighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses (Q21=1)
    mutate(Type = "Unweighted") |>
    select(Type, educ, percentage),
  
  # Weighted data - 
  weighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses
    mutate(
      Type = "Weighted",
      percentage = weighted_percentage
    ) |>
    select(Type, educ, percentage)
)

# Create a grouped bar chart
ggplot(educ_viz_data, 
       aes(x = educ, y = percentage, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(
    title = "Weighted vs. Unweighted 'Yes' Responses by Education",
    subtitle = "Q21: Do you plan to vote in the November election?",
    y = "Percentage (%)",
    x = "Education Level"
  ) +
  scale_fill_manual(values = c("Unweighted" = "#619CFF", "Weighted" = "#F8766D")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Task 10: Reflection Questions

Does the visualization make it easier to see the differences between weighted and unweighted results? Why or why not?

The visualizationd does make it easier to see the difference because the two categories are side by side. It is also helpful to see the numbers next to each other as opposed to on two separate charts. The graph also shows how close the other two groups are. 
## Task 11: Summary

Based on your analysis of weighted and unweighted results by education level, write a brief (2-3 paragraph) journalistic summary of what you found. You should try to address:

1. How education level relates to voting intentions
2. How weighting affected your understanding of this relationship
3. What this means for interpreting polling results in news reporting

Data shows that college-educated citizens are the more likely to vote than those with less education, with 93.5% of college-educated citizens reporting that they are voting in the upcoming election. 85.1% of those with some college education reported that they will vote, along with 70.8% of those with a high school or less education. Those with a high school or less education were the most unsure about whether or not they will vote in the upcoming election, with 13.2% responding unsure. 

Weighting the data helped me to get a better sense of how the high school or less group responded to the poll, as they were under-represented in the unweighted data. Since the data barely affected the other two subgroups, that shows that those groups were adequately represented in the sample. When interpreting polls in news reporting, reporters should be sure to double check the methodology of the poll to see that the population demographics are accurately represented in the sample. This will bring the data as close as possible to the true beliefs of the public. 
## Task 12: Final Reflection Questions

1. Why is it important for journalists to understand survey weighting when reporting on polls?

Journalists should understand survey weighting because we must understand the methods of a survey in order to ensure it is accurate. It can also help us to word stories about polls so we make sure we are not saying the poll is a definite predictor of what will happen. It is also important for understanding why poll results differ. 

2. How might the differences between weighted and unweighted results affect how you would report on this data?

Whether one chooses to report on the weighted or unweighted data will affect what numbers they share in their story. The differences would also dictate how the story is worded, such as mentioning that the results are/are not reflective of the true demographic's population, and how the poll cannot predict the future. 

3. What additional information would you want to know about how the weights were calculated before using this data in a news story?

I would want to know what percentages they are basing the weight off of and if they are reflective of the most recent data. I would also want to know the method of their weighting (how they apply it) and if that system has been trustworthy in the past. 
