---
title: "final_project"
output: html_document
---
Week 13 works (Final project: trials, errors and lessons)

Questions: 
I couldn't figure out what I was doing wrong with the T test. I tried to copy from the ANOVA activity, but it kept saying "no factors in the fitted model."

How do I do a "correlation test" as mentioned in your feedback?
```{r}
library(tidyverse)
library(kableExtra)

emissions <- read_csv("https://raw.githubusercontent.com/npilsbury/jour405_spring25/refs/heads/main/project/Emissions.csv")

wins <- read_csv("https://raw.githubusercontent.com/npilsbury/jour405_spring25/refs/heads/main/project/Project%20Data%20-%20Sheet1.csv")

#Summary

emissions_summary <- emissions %>% mutate(mean_emissions2023 = mean(Emissions2023), median_emissions2023= median(Emissions2023), mean_emissions2024= mean(Emissions2024), median_emissions2024 = median(Emissions2024), sd_emissions2023 = sd(Emissions2023), sd_emissions2024 = sd(Emissions2024))

wins_summary <- wins %>% mutate(mean_wins2023= mean(Wins2023), mean_wins2024 = mean(Wins2024), median_wins2023 = mean(Wins2023), median_wins2024 = median(Wins2024), sd_wins2023 = sd(Wins2023), sd_Wins2024 = sd(Wins2024))


```
```{r}
#ANOVA

anova_result_wins <- aov(Wins2023 ~ Wins2024, data= wins)
summary(anova_result_wins)

anova_result_emissions <- aov(Emissions2023 ~ Emissions2024, data= emissions)
summary(anova_result_emissions)
```
```{r}
#Z-Score
emissions_summary <- emissions_summary %>% mutate(z_score = (Emissions2023 - mean_emissions2023) / sd_emissions2023)

emissions_summary |>
  select(Team, Emissions2023, z_score) |>
  head(10) |>
  kable(digits = 2)

emissions_summary <- emissions_summary %>% mutate(z_score = (Emissions2024 - mean_emissions2024) / sd_emissions2024)

emissions_summary |>
  select(Team, Emissions2024, z_score) |>
  head(10) |>
  kable(digits = 2)

wins_summary <- wins_summary %>% mutate(z_score = (Wins2023 - mean_wins2023) / sd_wins2023)

wins_summary |>
  select(Team, Wins2023, z_score) |>
  head(10) |>
  kable(digits = 2)

wins_summary <- wins_summary %>% mutate(z_score = (Wins2024 - mean_wins2024) / sd_Wins2024)

wins_summary |>
  select(Team, Wins2024, z_score) |>
  head(10) |>
  kable(digits = 2)

```
```{r}
#T test (i do not know what to do)
tukey_result_wins <- TukeyHSD(anova_result_wins)
tukey_result_emissions <- TukeyHSD(anova_result_emissions)
print(tukey_result_emissions)
```
```{r}
#Linear Regression
data <- merge(emissions, wins)
data %>% 
  ggplot(aes(x = `Emissions2023`, y = `Wins2023`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Emissions 2023", y = "Wins 2023", title = "Scatterplot of Emissions vs Wins")

model2023 <- lm(`Wins2023` ~ `Emissions2023`, data = data)
summary(model2023)

data %>% 
  ggplot(aes(x = `Emissions2024`, y = `Wins2024`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Emissions 2024", y = "Wins 2024", title = "Scatterplot of Emissions vs Wins")

model2024 <- lm(`Wins2024` ~ `Emissions2024`, data = data)
summary(model2024)

```


Week 12 Work

```{r}
library(tidyverse)
```

```{r}
Emissions <- read_csv("https://raw.githubusercontent.com/npilsbury/jour405_spring25/refs/heads/main/project/Emissions.csv")
View(Emissions)

Wins <- read_csv("https://raw.githubusercontent.com/npilsbury/jour405_spring25/refs/heads/main/project/Project%20Data%20-%20Sheet1.csv")
View(Wins)
```

Types of Tests to Run

ANOVA: Used to test if there are real patterns in the data, or if it is just coincidental. Most of the teams emitted more carbon from travel during 2024, and 11 of the teams did worse, or had less wins in the 2024 season. ANOVA can test if these two factors are related or if it just ended up that way. 

Z-Score: Calculate the number of standard deviations something is from the mean. I think that this could be used for both the emissions data and the wins data. I can calculate the average amount of carbon emissions per year, as well as the average increase. I can then use z-scores to see if any university had an unusual change in the amount of emissions. I can do something similar for wins. 

Tukey's Test (?): This test might be useful to see which of the two colleges differ the most for the increase/decrease of both emissions and wins.

Other Analyses

I think that making charts and graphs to visualize the data would be helpful for comparing. I think a bar chart would be best.

Past Work

Project Details

Part 1: Comparing Big 10 Teams Changes Carbon Emissions 
- Individual teams increase/decrease of carbon emissions
- Comparing individual teams for 2023 and 2024
- Comparing individual teams changes from 2023 to 2024

Part 2: Big 10 Teams Carbon Emissions and Amount of Wins
- Are carbon emissions and the amount of wins a team have related?
- Do the teams that emit the most carbon emissions have the most wins? The least wins?

Data

Carbon Emissions: https://cnsmaryland.org/2024/12/11/big-ten-football-travel-carbon-emissions-expansion/ (also in project folder in CSV format)

Wins:
2024: https://bigten.org/fb/standings/2024/ (pdf in folder)
2023: https://bigten.org/fb/standings/2023/ (pdf in folder)
UCLA 2023:https://www.espn.com/college-football/team/schedule/_/id/26/season/2023
USC 2023: https://www.espn.com/college-football/team/schedule/_/name/usc/season/2023 
Oregon 2023: https://www.espn.com/college-football/team/schedule/_/id/2483/season/2023 
Washington 2023: https://www.espn.com/college-football/team/schedule/_/id/264/season/2023 
Compiled 2023 data to account for addition of 4 teams: https://docs.google.com/spreadsheets/d/1F5T3dpv0nkBDtk-4UeIJzp1RGHnu8ow7pXMRG40-Z0w/edit?usp=sharing (also pdf in folder)

Assignment Questions

The data is from 2023 and 2024. It focuses on regular season conference games for Big 10 football teams and their carbon emissions. The Big 10 added four teams for the 2024 season, all of which are on the West Coast. 

Both columns in the carbon emissions dataset are important. As for the wins data, only the conference game records are important, as they show how many wins each team had. 

As of now, the only question I have is how to import the data I have. I have both the CSV and the google sheet from 2023 pdf in the folder, but I am not sure how to load it.
