---
title: "Final_Project_Complete"
output: html_document
---
Here I am loading the libraries and the data needed for my analysis.
```{r}
library(tidyverse)
library(kableExtra)

emissions <- read_csv("https://raw.githubusercontent.com/npilsbury/jour405_spring25/refs/heads/main/project/Emissions.csv")

wins <- read_csv("https://raw.githubusercontent.com/npilsbury/jour405_spring25/refs/heads/main/project/Project%20Data%20-%20Sheet1.csv")
```
My data consists of Big Ten football team's carbon emissions and wins from 2023 and 2024. Since more teams joined the Big Ten in 2024, I added in their emissions and wins from the 2023 season, even though they were not in the Big Ten yet. The goal of my project is to determine if carbon emissions has any sort of impact on how many wins a team has in a season. 

Here I am using the "summary" feature to find the mean, median and standard deviation for the 2023 and 2024 emissions and wins for Big Ten football teams. This will be helpful for future calculations, as well as making predictions about if any team is an outlier in terms of emissions or wins. I am also combing the data into one dataset for future tests.

```{r}
emissions_summary <- emissions %>% mutate(mean_emissions2023 = mean(Emissions2023), median_emissions2023= median(Emissions2023), mean_emissions2024= mean(Emissions2024), median_emissions2024 = median(Emissions2024), sd_emissions2023 = sd(Emissions2023), sd_emissions2024 = sd(Emissions2024))

wins_summary <- wins %>% mutate(mean_wins2023= mean(Wins2023), mean_wins2024 = mean(Wins2024), median_wins2023 = mean(Wins2023), median_wins2024 = median(Wins2024), sd_wins2023 = sd(Wins2023), sd_Wins2024 = sd(Wins2024))

combined <- wins |> inner_join(emissions, join_by(Team))
```
The mean amount of carbon emissions for the 2023 season was 115,276 kilograms, while the mean for the 2024 season was 248,846.4 kilograms. The increase of emissions more than doubled from the 2023 to 2024 season. The median emissions of the 2023 season also nearly doubled in the 2024 season. The standard deviation more than tripled from the 2023 to 2024 season.
There was a 1.1 decrease in wins comparing the 2023 and 2024 season. The standard deviation also decreased by about 0.8. 

Calculating Z-Scores

I decided to calculate Z-scores to determine if there were any outliers in the data, or if anything was considered unusual. I also wanted to see if any university had a notable increase or decrease in emissions or wins when comparing the 2023 to 2024 year. I thought it would be interesting to see if the emissions and wins z-scores would be similar for each university. 

```{r}
emissions_summary <- emissions_summary |> mutate(z_score = (Emissions2023 - mean_emissions2023) / sd_emissions2023)

emissions_summary |>
  select(Team, Emissions2023, z_score) |>
  head(18) |>
  kable(digits = 2)

emissions_summary <- emissions_summary %>% mutate(z_score = (Emissions2024 - mean_emissions2024) / sd_emissions2024)

emissions_summary |>
  select(Team, Emissions2024, z_score) |>
  head(18) |>
  kable(digits = 2)

wins_summary <- wins_summary %>% mutate(z_score = (Wins2023 - mean_wins2023) / sd_wins2023)

wins_summary |>
  select(Team, Wins2023, z_score) |>
  head(18) |>
  kable(digits = 2)

wins_summary <- wins_summary %>% mutate(z_score = (Wins2024 - mean_wins2024) / sd_Wins2024)

wins_summary |>
  select(Team, Wins2024, z_score) |>
  head(18) |>
  kable(digits = 2)
```
Only two z-scores came up unusual-with both being for Washington. Washington had an unusual score of 2.20 for the 2024 emissions category, and an unusual score of 2.27 for the 2023 wins category. I thought it was interesting that the same team had an unusual score twice, but for different categories in different years. For 2023 emissions, Washington had a z-score of 1.99, which is on the brink of unusual. For 2024 wins, Washington had a z-score of -0.20. 
The 2024 emissions z-score analysis stuck out to me because there were 12 negative z-scores and 6 positive z-scores. This is interesting because it means that the teams with positive z-scores significantly brought up the mean, leaving a large gap between the mean and the other team's emissions. For this category, the highest z-score was Washington with 2.20, and the lowest was Rutgers with 0.11.
Another notable z-score was the -1.82 score for Purdue's 2024 wins. When looking at the raw data, Purdue is the only team to have decreased their emissions from the 2023 to 2024 season, so it is interesting that they had the largest z-score for the 2024 wins category. Conversely, Oregon had a z-score of 1.82 for 2024 wins and had higher emissions that other teams, with a z-score of 1.25 for 2024 emissions. 

ANOVA

ANOVA can be used to test if there are real patterns in the data, or if the data happens to be coincidental. When looking at the raw data, 17/18 of the teams increased their carbon emissions from the 2023 to 2024 season, and 11 of the teams had less wins in the 2024 season. I ran this test for both emissions and wins to see if there were patterns in the data and if 2023 wins can predict 2024 wins. 

```{r}
wins <- wins %>% 
  mutate(
    wins2023_category = cut(Wins2023, 
                          breaks = c(-Inf, 3, 7, Inf), 
                          labels = c("Low", "Medium", "High")),
    wins2024_category = cut(Wins2024, 
                          breaks = c(-Inf, 3, 7, Inf), 
                          labels = c("Low", "Medium", "High"))
  )
anova_result_wins <- aov(Wins2024 ~ wins2023_category, data = wins)
summary(anova_result_wins)

```
The F value is small, which means that there is more within group variance than between group variance. The F value is not too insignificant, as the values of wins range from 0-13, but it is still not large enough of a number to determine that 2023 wins can predict 2024 wins. The p-value is also 0.287, which means that the results are not statistically significant. 

```{r}
emissions <- emissions %>% 
  mutate(
    emissions2023_category = cut(Emissions2023, 
                          breaks = c(-Inf, 177600, 261700, Inf), 
                          labels = c("Low", "Medium", "High")),
   emissions2024_category = cut(Emissions2024, 
                          breaks = c(-Inf,177600, 261700, Inf), 
                          labels = c("Low", "Medium", "High"))
  )
anova_result_emissions <- aov(Emissions2024 ~ emissions2023_category, data = emissions)
summary(anova_result_emissions)
```
This ANOVA test resulted in an extremely low F value of 3.439. This is a very small number, especially when considering that the emissions hundreds of thousands of kilograms. This implies that there is more within group variance than between group variance - which also implies that the 2023 emissions cannot predict 2024 emissions. 
The p-value is also higher than 0.05, which means that the results are not statistically significant. 

Both ANOVA tests resulted in a low F value and a high p-value. Both tests revealed that there was more within group variance than between group variance, meaning that the 2023 values could not predict the 2024 values. 

Tukey Test

I decided to run a Tukey test to see how high, medium, and low emissions and wins differed for each time. I split the three categories up evenly based on the raw data. This test will show, on average, how much higher one category is than another. 

```{r}
TukeyHSD(anova_result_wins)
```
It is interesting that the high-low and high-medium comparisons both have a difference of about 2 wins. I expected a bigger difference, though the similarity makes sense when looking at the slight difference between the medium and low categories. This may be because there are less teams that fall into the high category, and many more teams fall within low or medium. In both years, the high category was only recorded 7 times, while the low category was recorded 14 times and medium was recorded 15 times. This test helped to identify that the medium and low categories were closer together since the high value was recorded about half as much as the other two. It also showed that, on average, the categories are fairly close together. 
All three p-values are above 0.05, meaning that the results are not statistically significant. 

```{r}
TukeyHSD(anova_result_emissions)
```
This had an interesting result, as it only showed the medium and low categories. This is because in the 2023 emissions category, there are no values high enough to reach the high category - all of the high values are in the 2024 category. While the 2024 category has an almost even split of high, medium and low, the 2023 category has 16 lows and 2 mediums. This shows how high the increases were in 2024- those emisisons added the whole extra top third of the dataset.
That being said, there is a fairly significant difference of 153,943.5 between the two categories, especially since the emissions are recorded in hundreds of thousands of kilograms. The p-value is slightly too high to be considered statistically significant, though better than past tests have shown. 

Both Tukey tests showed patterns in the data, specifically pertaining to oddities with how the high category was distributed. 

Linear Regression

I chose to do a linear regression model to visually determine if there is a linear relationship between a team's emissions and wins. The visual graph will help to place any outliers, or if there is a correlation between the two categories. The linear regression model will also provide an adjusted r-squared value, will help to determine how much of the independent variable is responsible for the dependent variable. Therefore, the adjusted r-squared will help me determine if there is a relationship between emissions and wins.
```{r}
combined %>% 
  ggplot(aes(x = `Emissions2023`, y = `Wins2023`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Emissions 2023", y = "Wins 2023", title = "Scatterplot of Emissions vs Wins 2023")

model2023 <- lm(`Wins2023` ~ `Emissions2023`, data = combined)
summary(model2023)

combined %>% 
  ggplot(aes(x = `Emissions2024`, y = `Wins2024`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Emissions 2024", y = "Wins 2024", title = "Scatterplot of Emissions vs Wins 2024")

model2024 <- lm(`Wins2024` ~ `Emissions2024`, data = combined)
summary(model2024)
```
Both of the scatterplots show a very very weak, positive correlation. 
The 2023 plot does not have many plot points that are near the line of the best fit, and barely has a positive correlation since the plot points are so spread out across the graph. The outlier in the upper right corner is very significant since it has the largest visual residual on the plot. The adjusted r-squared value for the 2023 plot is 10.14%. This means that only about 10% of the data can be reflected by this model. This model also has a p-value of 0.107, which is above the statistically significant p-value of 0.05.
The 2024 plot has data points that are slightly more clustered around the line of best fit, and visually does not have as high of residuals as the last graph. Despite this, the points are still very spread out and do not cluster around the line of best fit enough to determine a correlation. The line of best fit has a weaker positive line of best fit compared to the line of best fit. The adjusted r-squared is -2.546%, which means that 
the emissions provides no predictive value for determining wins and makes the model worse. To add on, the p-value is 0.4582, which means the data is not statistically significant. 
To conclude, the models and the figures that arose from them show that there is no relationship between the two variables. 

Pitch idea and reporting plan

Pitch: I think a story pitch based on this data could be in the context of four teams joining the Big Ten conference in 2024. Since the four teams that joined are on the West Coast, teams are letting out more carbon emissions from travel. Carbon emissions harm not only the planet, but can also threaten human health through worsening air quality, threats to food and water quality, and the overall warmth of the planet (EPA). I think this story can focus on how even though expanding the conference has negative implications for environment and health, it does not impact the performance or wins of a team. Though people may not think this is directly correlated, it could draw attention to the negative impacts of carbon emissions and how if they continue to increase, the health of athletes (and everyone else) will decrease - and then would ultimately actually impact wins. This would be an interesting story to follow over time as carbon emissions are expected to worsen, but may not have enough data for right now. 
The headline would read something like "Increased carbon emissions have no impact on Big Ten football wins, despite human health risks"

Potential Problems: A potential problem would be waiting a few years for more data. If carbon emissions do end up having enough significant impacts on health over the next few years, a reporter would need to wait to acquire that data. Another problem may be finding coaches and Big Ten staff who are willing to speak on the increase of carbon emissions due to travel. They may decline to comment, or give a general statement that does not add much to the story. It also may be hard to explain the lack of correlation in terms a general audience would understand. 

Reporting Plan: I would first acquire more data from over the years to have a well rounded and extensive report. I would then speak to climate change and EPA experts to analyze how carbon emissions ca impact human health, and if these effects are currently affecting us. I would also speak to Big Ten coaches and players to analyze any notable differences in air quality or access to healthy food as the warming planet changes availability and prices of goods - or any other health changes they may have noticed over the years. This can also include if warming temperatures make it more difficult to play, especially as they travel to warmer locations. I would also most likely have to get a general statement from the overall Big Ten. 
